\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{listings}

\lstdefinelanguage{Soda}{
    morekeywords={lambda, any, def, if, then, else, match, case, class, extends, abstract, end, inductive, this, subtype, supertype, false, true, not, and, or, package, import, theorem, directive, @tailrec, @override, @new},
    sensitive=true,
    morecomment=[s]{/*}{*/},
   morestring=[b]"
}

\lstset{frame=tb,
    language=Soda,
    aboveskip=3mm,
    belowskip=3mm,
    showstringspaces=false,
    columns=flexible,
    basicstyle={\small\ttfamily},
    numbers=none,
    numberstyle=\tiny\color{gray},
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{teal},
    breaklines=true,
    breakatwhitespace=true,
    tabsize=3
}

\begin{document}

\begin{lstlisting}
package soda.manual


\end{lstlisting}

This is a Soda tutorial written in Soda.
 Copyright 2020--2025 Julian Alfredo Mendez
 Version: 2025-03-28


\begin{lstlisting}



\end{lstlisting}

This tutorial is itself a "Hello world!" program.
 The piece of code that prints the message is at the end of the file.


\begin{lstlisting}



\end{lstlisting}

Source code written in Soda is intended to be descriptive and readable.
 It is usually written in different files, and each file has 'blocks'.
 These blocks are pieces of code that have some meaning.
 A block should be short, maybe less than 10 lines.
 However, it is more important to make things clear than concise.
 These are examples of blocks:
 1. a constant or function definition
 2. the beginning of a class definition
 3. the end of a class definition
 4. a block declaration of abstract constants and functions
 5. a definition of an algebraic data type
 6. a block of imports
 7. a package declaration
 8. a class alias
 9. a comment


\begin{lstlisting}



\end{lstlisting}

To declare a class, just add the reserved word `class` before a class name, and end it with
 the reserved word `end`.
 The reserved word `abstract` explicitly states whether constants and functions are required
 to instantiate a class. In the example below, no constants or functions are required.
 For the class name, it is recommended to use camel case style starting with a capital letter.
 The name could be a noun phrase or an adjective, but it should not be a verb.


\begin{lstlisting}


class Shape

  abstract

end


\end{lstlisting}

The reserved word `class` declares a type, a namespace, and a default constructor.
 It is possible to instantiate a class using the function `mk`. The notation looks
 as a static function of the type. For example, for `Shape` is `Shape .mk` without
 parentheses.


\begin{lstlisting}


class Movable

  abstract

end


\end{lstlisting}

It is recommended to indent the constants and functions declared inside.


\begin{lstlisting}


class EqualsExample


\end{lstlisting}

  Note that `abstract` does not contain any constants or functions in this example, and the
   declared constants and functions are in a different block.


\begin{lstlisting}


  abstract


\end{lstlisting}

  A constant does not have parameters and it is declared with the equals sign (`=`).
   For the constant name, it is recommended to use snake case and start in lowercase.
   The constant name should be a noun phrase.


\begin{lstlisting}


  my_number : Int = 2


\end{lstlisting}

  A function has parameters and a type. Functions, even with empty parameters, are evaluated
   every time they are invoked. The standard way of declaring and invoking a function
   with multiple parameters is with parameters separated by spaces and not by commas.
   For example, use `f (x) (y)` instead of `f(x, y)`.


\begin{lstlisting}


  f (x : Int) (y : Int) : Int = 2 * x + y


\end{lstlisting}

  Constants are only evaluated once, which is the first time they are needed.


\begin{lstlisting}


  first_result : Int = f (12) (4)


\end{lstlisting}

  In a function call, the parameters can be specified with the colon-equals sign (`:=`).
   This is especially recommended when several parameters are of the same type.


\begin{lstlisting}


  second_result : Int = f (x := 20) (y := -10)


\end{lstlisting}

  It is allowed to use the optional reserved word `def` to define constants and functions.


\begin{lstlisting}


  def a_constant : Int = 1

  def a_function (x : Int) : Int = x * x

end


\end{lstlisting}

It is possible to define algebraic data types, especially inductive data types.
 The reserved word for that definition is `inductive` with the name of the data type.
 The constructors are defined after the line with the `inductive` reserved word.
 Each constructor takes a possibly empty number of parameters and returns an element of the type being defined.
 Each parameter is separated by the type arrow (`->`) and the same is used as a result of the construction.
 It is mandatory to give a name to each parameter.



\begin{lstlisting}


inductive NaturalNumber
  Zero : NaturalNumber
  Successor : (n : NaturalNumber) -> NaturalNumber


\end{lstlisting}

A class can extend other classes by using the reserved word `extends`.
 Abstract classes contain a block with the reserved word `abstract`, and they are the only
 classes that can be extended.
 By contrast, concrete classes are declared with parentheses `(` and `)` and cannot be
 extended. Algebraic data types cannot be extended either.


\begin{lstlisting}


class RegisteredPerson


\end{lstlisting}

  The block starting with `abstract` denotes constants or functions that need to be defined
   in extending classes. Only one `abstract` block should be defined per class, without
   leaving lines between the declared attributes.


\begin{lstlisting}


  abstract
    first_name : String
    last_name : String


\end{lstlisting}

  If a constant or function is not meant to be exported, its name should start with an
   underscore.


\begin{lstlisting}


  _separator = " "


\end{lstlisting}

  Strings can be concatenated by using the plus sign (`+`).


\begin{lstlisting}


  full_name = first_name + _separator + last_name

end

class Agent

  abstract
    identifier : String

end


\end{lstlisting}

A concrete class needs as parameters all the constants and functions that have not been
 defined in its super classes. Observe that `extends` has to be in the same block as
 `class`, which need to be in a different block from `abstract`.


\begin{lstlisting}


class RegisteredPersonAgent
  extends
    Agent
    RegisteredPerson

  abstract
    identifier : String
    first_name : String
    last_name : String

end

class Element


\end{lstlisting}

  In this class, `accept` is a function that takes an object of type `Visitor` and returns
   an object of type `Boolean`. This is indicated with the type arrow `->`.


\begin{lstlisting}


  abstract
    accept : Visitor -> Boolean

end

class Visitor


\end{lstlisting}

  In this class, `visit` is a function defined from `Element` to `Boolean`.


\begin{lstlisting}


  abstract
    visit : Element -> Boolean

end

class Item
  extends Element

  abstract
    identifier : Int


\end{lstlisting}

  It is possible to refer to an object instance by using reserved word `this`.
   The dot (`.`) notation is the standard way of accessing attributes and methods of an
   object. The space before the dot is to improve readability and it is not necessary.
   Lambda functions are declared with the reserved word `lambda` and a long arrow (`-->`).
   Notice the difference between the type arrow (`->`) and the lambda arrow (`-->`).


\begin{lstlisting}


  accept : Visitor -> Boolean =
    lambda visitor -->
      visitor .visit (this)

end

class PersonName

  abstract
    name : String


\end{lstlisting}

  It is possible to override a function by using the `@override` annotation.
   This is intended only for exceptional cases, like the `toString` function, or a
   diamond-shaped class hierarchy.


\begin{lstlisting}


  @override
  toString = name

end


\end{lstlisting}

A class can be parameterized using square brackets ('[' and ']').
 The class parameter needs to be of type Type.


\begin{lstlisting}


class MyList [A : Type]

  abstract

end


\end{lstlisting}

It is possible to have multiple type parameters.


\begin{lstlisting}


class MyPair [A : Type] [B : Type]

  abstract
    fst : A
    snd : B

end


\end{lstlisting}

It is possible to define a pair as a data type.


\begin{lstlisting}


inductive AnotherPair [A : Type] [B : Type]
  AnotherPair_ : (fst : A) -> (snd : B) -> AnotherPair [A] [B]


\end{lstlisting}

It is possible to define a triple as a data type.


\begin{lstlisting}


inductive Triple [A : Type] [B : Type] [C : Type]
  Triple_ : (fst : A) -> (snd : B) -> (trd : C) -> Triple [A] [B] [C]


\end{lstlisting}

The parameter type can be constrained using `subtype` and `supertype`.
 In that case, it is not necessary to declare the parameter to be of type Type.
 For example, `A subtype B` means that `A` is of type Type and it is a subtype of `B`.


\begin{lstlisting}


class ShapePainter [A subtype Shape]

  abstract

end


\end{lstlisting}

The following class contains examples given as pieces of code.


\begin{lstlisting}


class Manual

  abstract


\end{lstlisting}

  The first line in this file is the package declaration. It contains the reserved word
   `package` followed by the package name. The recommended package naming convention is to
   start with `soda.`, which helps to avoid name conflicts when it is translated to Scala.
   The package declaration usually goes in a separate file called `Package.soda`.


\begin{lstlisting}



\end{lstlisting}

  It is possible to import classes by listing them under the reserved word `import`.
   Imported classes can also be declared in the `Package.soda` file, when they are global
   for the whole package. The list of imported classes can be used to detect and control which
   classes may be producing side effects.


\begin{lstlisting}


  import
    java.util.Date

  a = 1

  b : Int = 2


\end{lstlisting}

  An instance of a JVM class can be created with the `@new` annotation. If the code is
   translated to Scala 3, this annotation is not required.


\begin{lstlisting}


  now : Date = @new Date ()

  plus_one (x : Int) : Int = x + 1


\end{lstlisting}

  A piecewise function can be defined using an `if`-`then`-`else` structure. The condition in
   the `if` is evaluated, and then only the corresponding branch is evaluated.


\begin{lstlisting}


  max (x : Int) (y : Int) : Int =
    if x > y
    then x
    else y


\end{lstlisting}

  Scala sequences (`Seq`) can be used, as well as other basic Scala classes.


\begin{lstlisting}


  plus_one (sequence : Seq [Int] ) : Seq [Int] =
    sequence .map (lambda element --> element + 1)


\end{lstlisting}

  A synonym for `lambda` is `any`, which sometimes brings more readability.


\begin{lstlisting}


  plus_two (sequence : Seq [Int] ) : Seq [Int] =
    sequence .map (any element --> element + 2)


\end{lstlisting}

  Boolean values `false` and `true` are available.


\begin{lstlisting}


  my_not (x : Boolean) : Boolean =
    if x
    then false
    else true

  my_and (x : Boolean) (y : Boolean) : Boolean =
    if x
    then y
    else false

  my_or (x : Boolean) (y : Boolean) : Boolean =
    if x
    then true
    else y


\end{lstlisting}

  Boolean values have the standard `not`-`and`-`or` functions.


\begin{lstlisting}


  my_xor (x : Boolean) (y : Boolean) : Boolean =
    (x or y) and not (x and y)


\end{lstlisting}

  It is possible to use pattern matching with the reserved words `match` and `case`.
   The result of the matching case is put after a long double arrow `==>`.
   The order matters, so the remaining cases are capture by the last variable.
   Please notice the difference between the case arrow (`==>`), the lambda arrow (`-->`), and
   the type arrow (`->`).


\begin{lstlisting}


  if_then_else [A : Type] (condition : Boolean) (if_true : A) (if_false : A) : A =
    match condition
      case true ==> if_true
      case false ==> if_false


\end{lstlisting}

  It is possible to use pattern matching to obtain the content of an object.
   In the case of pattern matching, the constructor name is the class name with an underscore
   as suffix.
   Unlike with `mk`, the parameters need to be retrieved all together in a tuple.


\begin{lstlisting}


  get_left [A : Type] [B : Type] (pair : MyPair [A] [B]) : A =
    match pair
      case MyPair_ (left , right) ==> left


\end{lstlisting}

  A constant or function name starting with underscore indicates that the constant or
   function is private, and therefore is not visible outside the class.


\begin{lstlisting}


   _my_private_function (x : Float) : Float =
     x * x + x + 1


\end{lstlisting}

  A tail recursive function cannot be declared inside another function, and its name should
   start with underscore. The annotation `@tailrec` helps to ensure that the tail recursion
   is detected and optimized when it is translated to Scala.


\begin{lstlisting}


  @tailrec
  _tailrec_sum (n : Int) (accum : Int) : Int =
    if n < 0
    then accum
    else _tailrec_sum (n - 1) (n + accum)

  sum (n : Int) =
    _tailrec_sum (n) (0)

end


\end{lstlisting}

The class `Fold` has a 'left fold', which is a functional approach to iterations. Starting
 with an initial value (`initial`), it traverses a sequence (`sequence`) applying a function
 (`next`) using the current element in the sequence and the result of the previous
 computation.


\begin{lstlisting}


class Fold

  abstract


\end{lstlisting}

  Note that the type parameters must be specified in the function call:
   `_tailrec_foldl [A] [B]`...
   The sequence constructor `+:` is defined by `Seq`.
   This is equivalent to the more common constructor `::`, when it is used for instances of
   `List`. `Nil` is the constructor for an empty `Seq`.


\begin{lstlisting}


  _tailrec_foldl [A : Type] [B : Type] (sequence : Seq [A] ) (current : B)
      (next : B -> A -> B) : B =
    match sequence
      case Nil ==> current
      case (head) +: (tail) ==>
        _tailrec_foldl [A] [B] (tail) (next (current) (head) ) (next)


\end{lstlisting}

  Ideally, each object should have one responsibility or purpose. The function `apply`
   defines the main responsibility of an object.


\begin{lstlisting}


  apply [A : Type] [B : Type] (sequence : Seq [A] ) (initial : B) (next : B -> A -> B) : B =
    _tailrec_foldl [A] [B] (sequence) (initial) (next)

end


\end{lstlisting}

A piece of code of the destination language can be included with the reserved word
 `directive`. In this example, we can define the concept of successor for integers for the
 translation to Scala.


\begin{lstlisting}


directive scala
object Succ_ {
  def unapply (n : Int) : Option [Int] =
    if (n <= 0) None else Some (n - 1)
}


\end{lstlisting}

The class `Range` generates a sequence of length `n` of consecutive natural numbers starting
 from 0.


\begin{lstlisting}


class Range

  abstract


\end{lstlisting}

  Note that `Range` processes any negative number as 0. This is done by `Succ_`, which only
   returns values for positive values of `n`.


\begin{lstlisting}


  _tailrec_range (non_negative_number : Int) (sequence : Seq [Int] ) : Seq [Int] =
    match non_negative_number
      case Succ_ (k) ==>
        _tailrec_range (k) ( (k) +: (sequence) )
      case _otherwise ==> sequence

  apply (length : Int) : Seq [Int] =
    _tailrec_range (length) (Nil)

end


\end{lstlisting}

`Factorial` shows an example of how to compute the function factorial using a left fold.


\begin{lstlisting}


class Factorial

  abstract

  fold = Fold .mk

  range = Range .mk

  apply (n : Int) : Int =
    fold .apply [Int] [Int] (range .apply (n) ) (1) (
      lambda accum -->
        lambda k --> (accum * (k + 1) ) )

end


\end{lstlisting}

The main class has to be named `Main` and requires a `main` function that receives an
 `Array [String]` and returns a `Unit`. Only one main class per package is allowed.


\begin{lstlisting}


class Main


\end{lstlisting}

  An output to the standard output can be sent with a `println` command.
   This is a shorter form of JVM's `System.out.println`.


\begin{lstlisting}


  main (arguments : Array [String] ) : Unit =
    println ("Hello world!")

end


\end{lstlisting}

The main class has an extending concrete class. The class that needs to be invoked in a
 translation to Scala is `EntryPoint`.


\begin{lstlisting}


\end{lstlisting}

\end{document}

